\documentclass[../main.tex]{subfiles}
\begin{document}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \section{Characteristics of Time Series}
    The study of time series is devoted to application of statistical techniques to data collected over multiple time steps. Traditional techniques are not useful here due to strong correlation between successive observations.\newline
    Two separate but not necessarily mutually exclusive approaches exist to study time series
    \begin{enumerate}
        \item Time Domain approach
        This approach gives importance to lagged relationships, such as how what happens today will effect the obervation 7 days later.
        \item Frequency Domain approach
        This approach gives importance to investigation ofo cycles, such as economic cycle of periods of expansion/recession.
    \end{enumerate}

    A time series is a collection of random variables $x_{1}, x_{2}, x_{3}, \ldots$ and can in general be denoted by $\{x_{t}\}$. Such a collection of random variables indexed by time is also called a \textbf{random process} and the values collected are often called the \textbf{realization} of the random process. In the context of time series analysis, the two can be used interchangeably.\newline

    The sampling rate can serve an important purpose when representing time series data. An insufficient sampling rate can completely change the appearance of the data. For instance, when recording the motion of a wheel, it can appear moving backwards at a low enough recording rate. This distortion is called \textbf{aliasing}.

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \subsection{Time Series Formulations}
    %%%%%%%%%%%%%%%%%%%%
    \subsubsection{White Noise}
    The simplest time series is just a series of uncorrelated variables. It has three flavours
    \begin{enumerate}
        \item white noise $w_{t} \sim wn(0, \sigma_{w}^{2})$
        \item white independent noise $w_{t} \sim iid(0, \sigma_{w}^{2})$
        \item Gaussian white noise $w_{t} \sim iid \mathcal{N}(0, \sigma_{w}^{2})$
    \end{enumerate}
    Just the noise is particularly difficult to model and we introduce some kind of correlation and smoothness into the model.

    %%%%%%%%%%%%%%%%%%%%
    \subsubsection{Moving Average and Filtering}
    Any time series can be smoothed by considering moving averages. A very simple example can be taking the average of three consecutive observations
    \begin{align*}
        v_{t} = \frac{1}{3}(x_{t-1} + x_{t} + x_{t+1})
    \end{align*}
    Care must be taken to ensure that the sum of the coefficients of all the terms is one.\newline

    As is evident, if we take more number of such terms, the slower oscillations become more prominent in comparison to the faster oscillations making moving averages work as a kind of \textbf{filter}.

    %%%%%%%%%%%%%%%%%%%%
    \subsubsection{Autoregression}
    This involves using a regression like function to predict the value of the time series at the current timestep. A simple example would be
    \begin{align*}
        x_{t} = x_{t-1} + 0.9x_{t-2} + w_{t}
    \end{align*}
    where $w_{t}$ is noise. Here we are using the observations from the last two timesteps to make our predictions. This idea can be extended to include more terms and even difference of terms as part of the equation. We make an assumption for some initial values to calculate $x_{1}$ and $x_{2}$.

    %%%%%%%%%%%%%%%%%%%%
    \subsubsection{Random Walk with Drift}
    The prediction function can be defined as
    \begin{align*}
        x_{t} &= \delta + x_{t-1} + w_{t}\\
        x_{t} &= \delta t + \sum_{t} w_{t}
    \end{align*}
    where $\delta$ is the drift. Setting $\delta = 0$ is same as a random walk.

    %%%%%%%%%%%%%%%%%%%%
    \subsubsection{Signal in Noise}
    Sometimes, the signal can be seen as having a periodic component with some noise
    \begin{align*}
        x_{t} = A\cos (2\pi \omega t + \phi) + w_{t}
    \end{align*}
    where $\phi$ is the phase or the offset of the signal, $\omega$ is the time period and $A$ is the amplitude. $w_{t}$ is noise that we cannot predict.\newline

    Let $\sigma_{w}$ denote the variance of the noise. Then, \textbf{signal-to-noise ratio} is the ratio of the amplitude of the signal to the variance of noise. The higher this quantity, the easier it is to detect the signal. In real data, the signal is always obscured by some noise.
    

\end{document}